---
output:
  pdf_document: default
  html_document: default
---
% C3W3 - Managing Data Frames with `dplyr` - Hugh Nguyen
%

```{r, echo=FALSE, results="hide"}
options(width = 50)
```

- arrange: re-order rows of a df

- filter: extract subset of rows of df BASED ON logical conditions

- select: return subset of cols of a df

- mutate: add new var or transform existing var in a df

- rename: rename cols/vars

- summarize: summarize statistics

- %>%: pipeline operation

# dplyr

The data frame is a key data structure in statistics and in R.

* There is one observation per row

* Each column represents a variable or measure or characteristic

* Primary implementation that you will use is the default R
  implementation

* Other implementations, particularly relational databases systems


# dplyr

* Developed by Hadley Wickham of RStudio

* An optimized and distilled version of `plyr` package (also by Hadley)

* Does not provide any "new" functionality per se, but **greatly**
  simplifies existing functionality in R

* Provides a "grammar" (in particular, verbs) for data manipulation

* Is **very** fast, as many key operations are coded in C++


# dplyr Verbs

* `select`: return a subset of the columns of a data frame

* `filter`: extract a subset of rows from a data frame based on
  logical conditions

* `arrange`: reorder rows of a data frame


* `rename`: rename variables in a data frame

* `mutate`: add new variables/columns or transform existing variables

* `summarise` / `summarize`: generate summary statistics of different
  variables in the data frame, possibly within strata

There is also a handy `print` method that prevents you from printing a
lot of data to the console.



# dplyr Properties

* The first argument is a data frame.

* The subsequent arguments describe what to do with it, and you can
  refer to columns in the data frame directly without using the $
  operator (just use the names).

* The result is a new data frame

* Data frames must be properly formatted and annotated for this to all
  be useful


# Load the `dplyr` package


This step is important!

```{r}
library(dplyr)
```


# `select`

```{r}
chicago <- readRDS("chicago.rds")
dim(chicago)
head(select(chicago, 1:5))
```


# `select`

```{r}
names(chicago)[1:3]
head(select(chicago, city:dptp))
```

# `select`

In dplyr you can do

```{r,eval=FALSE}
head(select(chicago, -(city:dptp)))
```

Equivalent base R

```{r,eval=FALSE}
i <- match("city", names(chicago))
j <- match("dptp", names(chicago))
head(chicago[, -(i:j)])
```



# `filter`

```{r}
chic.f <- filter(chicago, pm25tmean2 > 30)
head(select(chic.f, 1:3, pm25tmean2), 10)
```

# `filter`

```{r}
chic.f <- filter(chicago, pm25tmean2 > 30 & tmpd > 80)
head(select(chic.f, 1:3, pm25tmean2, tmpd), 10)
```


# `arrange`

Reordering rows of a data frame (while preserving corresponding order
of other columns) is normally a pain to do in R.

```{r}
chicago <- arrange(chicago, date)
head(select(chicago, date, pm25tmean2), 3)
tail(select(chicago, date, pm25tmean2), 3)
```

# `arrange`

Columns can be arranged in descending order too.

```{r}
chicago <- arrange(chicago, desc(date))
head(select(chicago, date, pm25tmean2), 3)
tail(select(chicago, date, pm25tmean2), 3)
```


# `rename`

Renaming a variable in a data frame in R is surprising hard to do!

```{r,tidy=FALSE}
head(chicago[, 1:5], 3)
chicago <- rename(chicago, dewpoint = dptp, 
                  pm25 = pm25tmean2)
head(chicago[, 1:5], 3)
```


# `mutate`

```{r, tidy=FALSE}
chicago <- mutate(chicago, 
                  pm25detrend=pm25-mean(pm25, na.rm=TRUE))
head(select(chicago, pm25, pm25detrend))
```

# `group_by`

Generating summary statistics by stratum

```{r, tidy=FALSE}
chicago <- mutate(chicago, 
                  tempcat = factor(1 * (tmpd > 80), 
                                   labels = c("cold", "hot")))
hotcold <- group_by(chicago, tempcat)
summarize(hotcold, pm25 = mean(pm25, na.rm = TRUE), 
          o3 = max(o3tmean2), 
          no2 = median(no2tmean2))
```


# `group_by`

Generating summary statistics by stratum

```{r, tidy=FALSE}
chicago <- mutate(chicago, 
                  year = as.POSIXlt(date)$year + 1900)
years <- group_by(chicago, year)
summarize(years, pm25 = mean(pm25, na.rm = TRUE), 
          o3 = max(o3tmean2, na.rm = TRUE), 
          no2 = median(no2tmean2, na.rm = TRUE))
```

```{r,echo=FALSE}
chicago$year <- NULL  ## Can't use mutate to create an existing variable
```


# `%>%`

```{r,tidy=FALSE,eval=FALSE}
chicago %>% mutate(month = as.POSIXlt(date)$mon + 1) 
	%>% group_by(month) 
	%>% summarize(pm25 = mean(pm25, na.rm = TRUE), 
          o3 = max(o3tmean2, na.rm = TRUE), 
          no2 = median(no2tmean2, na.rm = TRUE))
```

```{r,echo=FALSE}
chicago %>% mutate(month = as.POSIXlt(date)$mon + 1) %>% group_by(month) %>% 
summarize(pm25 = mean(pm25, na.rm = TRUE), o3 = max(o3tmean2, na.rm = TRUE), no2 = median(no2tmean2, na.rm = TRUE))

```


# dplyr

Once you learn the dplyr "grammar" there are a few additional benefits

* dplyr can work with other data frame "backends"

* `data.table` for large fast tables

* SQL interface for relational databases via the DBI package

# QUIZ

## Q1:
```{r}
download.file('https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2Fss06hid.csv'
              , 'ACS.csv'
              , method='curl' )

# Read data into data.frame
ACS <- read.csv('ACS.csv')

agricultureLogical <- ACS$ACR == 3 & ACS$AGS == 6
head(which(agricultureLogical), 3)
```

## Q2:

```{r}
library(jpeg)

# Download the file
download.file('https://d396qusza40orc.cloudfront.net/getdata%2Fjeff.jpg'
              , 'jeff.jpg'
              , mode='wb' )

# Read the image
picture <- jpeg::readJPEG('jeff.jpg'
                          , native=TRUE)

# Get Sample Quantiles corressponding to given prob
quantile(picture, probs = c(0.3, 0.8) )
```

## Q3:

```{r}
library("data.table")


# Download data and read FGDP data into data.table
FGDP <- data.table::fread('https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv'
                          , skip=4
                          , nrows = 190
                          , select = c(1, 2, 4, 5)
                          , col.names=c("CountryCode", "Rank", "Economy", "Total")
                          )

# Download data and read FGDP data into data.table
FEDSTATS_Country <- data.table::fread('https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv'
                                      )
                                      
mergedDT <- merge(FGDP, FEDSTATS_Country, by = 'CountryCode')

# How many of the IDs match?
nrow(mergedDT)
```

## Q4:

```{r}
# "High income: OECD" 
mergedDT[`Income Group` == "High income: OECD"
         , lapply(.SD, mean)
         , .SDcols = c("Rank")
         , by = "Income Group"]



# "High income: nonOECD"
mergedDT[`Income Group` == "High income: nonOECD"
         , lapply(.SD, mean)
         , .SDcols = c("Rank")
         , by = "Income Group"]

```

## Q5:

```{r}
library('dplyr')

breaks <- quantile(mergedDT[, Rank], probs = seq(0, 1, 0.2), na.rm = TRUE)
mergedDT$quantileGDP <- cut(mergedDT[, Rank], breaks = breaks)
mergedDT[`Income Group` == "Lower middle income", .N, by = c("Income Group", "quantileGDP")]
```
